---
title: "Final Project: Exploring the National Survey on Drug Use and Health "
author: "Suman Tripathy, Amanda Steigman"
date: "11/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(vcd)
library(forcats)
library(scales)
library(dplyr)
library(plyr)
```

## I. Introduction

Over the last few years, the opioid crisis in America has been worsening rapidly and is perhaps finally gaining the public and media awareness it needs The topic is both concerning and fascinating. When people think of drugs, the most obvious, widespread choices are alcohol and marijuana. The flashier ones that may come to mind afterwards include cocaine, LSD, ecstasy, and so on. 

Opioids, however, are getting more and more common. The National Institute on Drug Abuse found in January 2019 that "more than 130 people in the United States die after overdoing on opioids" every day. Opioids are most commonly prescription pain relievers but also include heroin and synthetic opioids such as fentanyl. The Centers for Disease Control and Prevention has released an estimate that opioid misuse in the US costs $78.5 billion a year in terms of healthcare costs, lost productivity, rehab treatments, and criminal justice activities.

An alarming number of doctors have been caught for inappropriate opiod prescription. Moreover, 21 to 29% of chronic pain patients have been found to misuse their prescribed opioids.  8 to 12% may develop an opioid use disorder, and 4 to 6% transition from an opioid prescripton to heroin use. Given these worrisome figures, we chose to explore data regarding the United States drug use and abuse. We are most interested in questions such as:

RANDOM BRAINSTORM THAT WE CAN PICK AND CHOOSE FROM / ADD TO
  - What are the interesting trends and rates of opioid abuse within different demographics?
  - What particular difficulties do youth who abuse opioids face?
  - Is there are strong relationship between opioid abuse and crime? What interesting depth can income and education add to this relationship?
  - What trends do we see between opioid abuse and other substance abuse?
  - Is crime at all connected with high opioid abuse, and how does this connection compare to crime rates tied to other drugs?
  - The cause of a lot of opioid misuse is physical illness, but what are differences between use, misuse, and abuse?  Are there any ties to mental illness?
  - What are the most common sources of obtaining opioids if not from a medical prescription?
  - 

## II. Data Sources

Our data source is the National Survey on Drug Use and Health from 2013. The primary investigators were the United States Department of Health and Human Services, the Substance Abuse and Mental Health Services Administration, and the Center for Behavioral Health Statistics and Quality. This survey has been conducted perioidically since 1971 and has transitioned from a traditional paper-pencil method to a computer-assisted interviewing method. It combines computer-assisted personal interviewing and computer-assisted self-interviewing. It is a highly private and confidential system in order to encourage participant honesty regarding these sensitive behaviors.

The demographic population is "the civilian, noninstituitionalized population of the United States ... who were 12 years of age or older at the time of the survey." Since 2002, NSDUH respondents have been incentivized with a $30 reward for taking the survey, and an improvement in the survey response rate was noted afterwards. This increase in response rate may also be tied to better public awareness about these critical issues.  

** SHOULD TRY TO LOOK FOR MORE CONCRETE INFO ON HOW THE SURVEY IS SENT OUT ** LOOK AT SAMPLE SECTION: https://www.icpsr.umich.edu/icpsrweb/NAHDAP/studies/35509/documentation# AND INCLUDE INFO ABOUT RESPONSE RATES

source: https://www.icpsr.umich.edu/icpsrweb/NAHDAP/studies/35509/documentation#

The dataset itself contains a myriad of informaition about the use of drugs, alcohol, and tobacco. It also covers mental and physical illnesses, population demographics, criminal activities, and more.  The target demographic is members of US households, age 12 and older. The loaded dataframe contains 55160 rows (participants) and 3142 columns (survey questions). Note that some of these columns are imputed values based on responses to other survey questions. 

As we are working with survey data, we will make sure to be aware of all the possible challenges that may come with this (for example, making sure that "missing" values are actually "missing" and are not logically "no.") We are checking this by going back to earlier survey sections to see how participants earlier responses may have caused parts of the survey to be skipped. We are also evaluating the imputed columns to see whether they are more cleaned and reliable than direct response columns.

## III. Data Transformation

We imported our data as a .rda file, which added a new challenge because loading and accessing the data was different than with the previous file types we have worked with. 

## IV. Missing Values

## V. Results



```{r}
mosaic(INCOME ~ SEX, 
       direction=c('v','h'), 
       labeling_args = list(rot_labels = c(0, 0, 0, 0), 
                            offset_varnames = c(left = 6), 
                            offset_labels = c(left = 2.6)),
       gp = grid::gpar(fill=c('#CCCCFF', '#CC99FF', '#9966CC', '#663399')),
       main = "Sex and Income of Oxycontin Users",
       drug_study_demographic,)
```

Above we can see that similar proportions of females and males have used Oxycontin non-medically (slightly more males than females). However, we do see some differences of use within gender when we vary income level. For both genders, we see that the highest proportion of users has an income of between $20,000 and $49,000. For females, usage drops off a bit after that (the majority of female users tend to be more poor). On the other hand, our male respondents with the highest income made up a larger portion of male users. When we look at income alone, we see a similar trend to male usage (about 30% of users make <$20,000 per year, 35% between $20,000 and $49,999, 15% between $50,000 and $74,999, and 20% make >$75,000 per year). 
## VI. Interactive Component

## VII. Conclusion

A major concern and limitation we recognized about the data is the inherent barriers to accurately assessing opioid misuse, most notably the term "misuse" itself. In the American Journal of Drug and Alcohol Abuse, Joseph Palamar wrote an article on this specific subject. The NSDUH 2013 survey acts questions about nonmedical use of drugs, "which it defined as when a prescription opioid was used when not presribed to the user or used for the experience or feeling caused by the drug." This term "nonmedical use" has inherent ambiguity, and recent studies of the NSDUH data estimate that about two-thirds of adults miusing opioids actually used the drug in order to relieve physical pain. Other adults used the medication to relieve stress or anxiety and sleep. Only about 11% of misusers are thought to actually use the opioid to "get high."  These figures are important to remember throughout our analysis of opioid users in order for us to understand the opioid crisis more precisely.

source: https://www.tandfonline.com/doi/full/10.1080/00952990.2018.1521826

To be consistent with many previous opioid investigations, we define frequent users as those with 100 or more days of nonmedical opioid use in the past year. Highly frequent users are those with 200 or more days of nonmedical use.
